Adaptive group-regularized logistic elastic net regression Supplementary material

### 2019-03-26 docs/supplement.Rnw
simulation 2 running

% set options
<<settings, include=FALSE, echo=FALSE>>=
opts_knit$set(root.dir="..", base.dir="../figs")
opts_chunk$set(fig.align='center', fig.path="../figs/", 
               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
@

% read figure code
<<figures, include=FALSE>>=
read_chunk("code/figures.R")
@

@@

  \subsection{Scenario 1}
	In this sparse scenario we assume differential signal between 5 groups of
	features. The average signal in a group increases by a factor 1.6 with group
	number:
	\begin{align*}
	  \bbeta & =  \begin{bmatrix} \smash[b]{\underbrace{ \begin{matrix} 
	  \mathbf{0} & \cdots & \mathbf{0} \end{matrix}}_{150}} 
	  & \smash[b]{\underbrace{ \begin{matrix} \beta_{1}^1 & \cdots & 
	  \beta_{50}^1 \end{matrix}}_{50}} & \cdots & 
	  \smash[b]{\underbrace{ \begin{matrix} \mathbf{0} & \cdots & \mathbf{0}
	  \end{matrix}}_{150}} & \smash[b]{\underbrace{ \begin{matrix} 
	  \beta_{1}^5 & \cdots & \beta^5_{50} \end{matrix}}_{50}} \end{bmatrix} \tr,
	  \\
	  & \, \\
	  \beta_{j}^g & \sim \mathcal{U} (0, b_1 \cdot1.6^{g-1}) \text{, } g=1, \dots,
	  5  \text{, } j=1, \dots, 50,
	\end{align*}
	where $\mathcal{U} (a, b)$ denotes the uniform distribution on $(a,b)$. The
	model parameters clearly do not follow the relatively heavy-tailed elastic net
	distribution. Because of this mismatch between the assumed and true 
	distributions, we expect poor performance by \texttt{gren}. Furthermore, we
	introduce correlation between the features by having covariance elements
	$\bSigma_{ij} = 0.5^{|i-j|}$.

	\subsection{Scenario 2}
	The second scenario contains many groups of features and is geared towards the
	non-adaptive group-regularized methods. These methods are able to incorporate
	the grouping of the features, but were originally designed for small 
	interpretable groups, such as dummies of a categorical variable. \texttt{gren}
	on the other hand, relies on large groups of features to accurately estimate
	group-specific penalty weights. The estimation of many penalty weights in
	small groups is inefficient and computationally challenging. We therefore
	expect \texttt{gren} to under-perform compared to the other methods.

	For this scenario we follow the second simulation example in 
	\cite{meier_group_2008}. They set $n=100$, $G=250$, and $p=1000$, such that we
	have 4 features per group. The first 10 groups are active and those 40 model
	parameters are set to 0.2. The predictors are simulated from a $p$-dimensional
	Gaussian distribution with covariance matrix $\bSigma_{ij} = 0.5^{|i-j|}$.

	\subsection{Scenario 3}
	A third scenario is modelled by considering a situation in which there is no
	differential signal between the groups, but the correlation within groups is
	very high. We conjecture that this leads to overestimation of effect within
	the groups by \texttt{gren} and consequently, performance will suffer.

	To that end we sample the predictors in $G=5$ separate blocks of 200 
	variables, corresponding to the partitioning. The variances of the features
	are set to one, while all covariances are 0.7. The 1000 model parameters are
	simulated in two steps. First, we draw 1000 parameters from the elastic net
	distribution, parametrised as in \cite{friedman_regularization_2010}, where
	$\alpha=0.5$ and $\lambda=100$. The second step consists of setting the 
	(in absolute value) smallest 100 parameters in a group to zero. This results
	in a total of 500 zero and 500 non-zero parameters, evenly distributed over
	the five groups.

	\subsection{Scenario 4}
	This scenario is similar to Scenario 3, except each groups penalty parameter
	is multiplied by one of : $\lambda' \in \{ 0.14, 0.51, 1.95, 7.39 \}$ and the
	covariance matrix of the predictors is block diagonal, with block sizes $25
	\times 25$, off-diagonals set to $\rho=0.7$, and diagonals to $\sigma^2 = 1$.
	We expect \texttt{gren} to pick up this differential signal and outperform the
	other methods.

	\subsection{Scenario 5}
	In the last scenario we create a sparse true model that contains some 
	differential signal. We impl

>>>>

% set options
<<settings, include=FALSE, echo=FALSE>>=
knitr::opts_knit$set(root.dir="..", base.dir="../figs")
knitr::opts_chunk$set(fig.align='center', fig.path="../figs/", 
                      echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
@

% read figure code
<<figures, include=FALSE>>=
knitr::read_chunk("code/figures.R")
@

@@

  \subsection{Scenario 1}
	In this sparse scenario we assume differential signal between 5 groups of
	features. The average signal in a group increases by a factor 1.6 with group
	number:
	\begin{align*}
	  \bbeta & =  \begin{bmatrix} \smash[b]{\underbrace{ \begin{matrix} 
	  \mathbf{0} & \cdots & \mathbf{0} \end{matrix}}_{150}} 
	  & \smash[b]{\underbrace{ \begin{matrix} \beta_{1}^1 & \cdots & 
	  \beta_{50}^1 \end{matrix}}_{50}} & \cdots & 
	  \smash[b]{\underbrace{ \begin{matrix} \mathbf{0} & \cdots & \mathbf{0}
	  \end{matrix}}_{150}} & \smash[b]{\underbrace{ \begin{matrix} 
	  \beta_{1}^5 & \cdots & \beta^5_{50} \end{matrix}}_{50}} \end{bmatrix} \tr,
	  \\
	  & \, \\
	  \beta_{j}^g & \sim \mathcal{U} (0, b_1 \cdot1.6^{g-1}) \text{, } g=1, \dots,
	  5  \text{, } j=1, \dots, 50,
	\end{align*}
	where $\mathcal{U} (a, b)$ denotes the uniform distribution on $(a,b)$. We set
	$b_1 \approx 0.18$ such that the mean expected signal 
	$\overline{\E(\beta_j^g)}=0.07$. Furthermore, we
	introduce correlation between the features by setting the covariances
	$\bSigma_{ij} = 0.5^{|i-j|}$.The
	model parameters clearly do not follow the relatively heavy-tailed elastic net
	distribution. Because of this mismatch between the assumed and true 
	distributions, we expect poor performance by \texttt{gren}.

	\subsection{Scenario 2}
	For this scenario we loosely follow the simulation setup in 
	\cite{meier_group_2008}. We create $G=300$ groups of features and $p=900$
	features in total, such that we
	have 3 features per group. There are 50 active groups, with 2 active 
	parameters with the same value and one inactive parameter per group, such 
	that in total we have 100 non-zero model parameters. The active model 
	parameters are fixed and decrease linearly from $30/51$ to $3/510$, such that
	the mean parameter $\overline{\beta}_j=1/30$. In this example, we included
	an intercept; $\beta_0=10$.

	The predictors are sampled in such a way that they represent dummy variables
	for categorical variables. We start with sampling 300 latent variables:
	$\mathbf{z} \sim \mathcal{N} (\mathbf{0}, \bSigma)$, with 
	$\bSigma_{ij} = 0.5^{|i-j|}$, $i,j=1, \dots, G$. These latent latent 
	variables are then transformed to dummies by:
	$$
	x_{ij} = 
	\begin{cases}
	  1 & \text{if } z_g \in (q_i, q_{i+1}), \\
	  0 & \text{otherwise},
	\end{cases}
	$$
	where the $q_i$ are the $0$, $1/3$, $2/3$, and $1$th quantiles of the normal
	distribution.

	The second scenario contains many groups of features and is geared towards the
	group lasso and its extensions. These methods are able to incorporate
	the grouping of the features, and were originally designed for small 
	interpretable groups, such as dummies of a categorical variable. \texttt{gren}
	on the other hand, relies on large groups of features to accurately estimate
	group-specific penalty weights. The estimation of many penalty weights in
	small groups is inefficient and computationally challenging. We therefore
	expect \texttt{gren} to under-perform compared to the other methods.

	\subsection{Scenario 3}
	In this scenario we sample the 1000 model parameters in $G=5$ groups of 200.
	For each group we generate the parameters in two steps. First we draw 200
	parameters from the elastic net distribution, parametrised as in 
	\cite{friedman_regularization_2010}, where $\alpha=0.5$ and $\lambda=100$.
	The second step consists of setting the 
	(in absolute value) smallest 100 parameters in a group to zero. This results
	in a total of 500 zero and 500 non-zero parameters, evenly distributed over
	the five groups. We sample the features from a multivariate normal 
	distribution with all variances 1 and all covariances 0.7.

	This third scenario is modelled by considering a situation in which there is no
	differential signal between the groups, but the correlation within groups is
	very high. We conjecture that this leads to unstable estimation of the 
	penalty multipliers by \texttt{gren} and consequently, performance will suffer.

	\subsection{Scenario 4}
	This scenario is similar to Scenario 3, except that we apply differential 
	shrinkage over the groups with penalty multipliers: 
	$\lambda' \in \{ 0.14, 0.51, 1.95, 7.39 \}$. 
	Furthermore, we change the covariance matrix of the predictors to a block
	diagonal, with block sizes $25
	\times 25$, such that each group of 200 features consists of 8 blocks of 
	correlated features. The covariance off-diagonals and diagonal in each block 
	are set to $\rho=0.7$ and $\sigma^2 = 1$, respectively.
	We expect \texttt{gren} to pick up this differential signal and outperform the
	other methods.

	\subsection{Scenario 5}
	In this last scenario, we create $G=10$ groups of 100 features.
	Of the 10 groups 8 do not contain any signal. In the two active groups, we set 
	85 model parameters to zero and 15 to 0.2 and 0.5, respectively, such that the
	mean signal $\overline{\beta}_j=21/2000$. The features are simulated from the
	same distribution as in Scenario 1.

	This sparse truth with differential signal between the groups is expected to
	yield a somewhat better performance by \texttt{gren} than the regular elastic
	net and \texttt{GRridge}. The group lasso extensions are expected to give a 
	similar performance.
